---
title: "Lesson 14 - Data Visualization in 2D (Live Coding)"
format: html
editor:
  markdown:
    wrap: 72
---

## Setup

We load the `tidyverse` (gives us `ggplot2`) and read in the cadet
data.

```{r}
# Load the tidyverse (gives us ggplot2)
library(tidyverse)

# Read in the cadet data
url <- "https://philipycho.github.io/public_usafa/DS220_Fall2026/simulated_cadets.csv"
# cadets <- read_csv("simulated_cadets.csv")
cadets <- read_csv(url)

# Take a first look at the variables and types
tibble(cadets)
```

## The three elements — a first scatterplot

Every plot needs **data** (which dataset), **aesthetics** (which
variable maps to which visual channel), and **geometry** (which type
of plot). A scatterplot uses `geom_point()` to show the relationship
between two *numerical* variables - it's the right choice here because
`avg_study_hours` and `gpa` are both numerical and we're asking about
a relationship, not the distribution of just one variable.

```{r}
# Task: build a scatterplot with avg_study_hours on x and gpa on y,
#       using cadets and geom_point()

# Task: in a comment, name the three elements of the plot you just
#       built - which dataset, which variable/channel, which geometry?

```

## Labeling a plot

Add a `labs()` layer (chained with `+`, not a pipe) to give the plot a
title and readable axis labels.

```{r}
# Task: copy your scatterplot from above and add a labs() layer with a
#       title, and x/y labels that are more readable than the raw
#       column names

```

## Fixing overplotting: transparency

With ~4,000 cadets, points pile on top of each other. Set `alpha`
(0-1) *inside* `geom_point()`, not inside `aes()`, to make points
partly transparent so overlap becomes visible as darker regions.

```{r}
# Task: copy your scatterplot and add alpha = 0.2 inside geom_point()
#       to reveal where points overlap

# Task: in a comment, describe the relationship using this vocabulary:
#       association? positive/negative? strong/weak? linear/non-linear?

```

## Fixing overplotting: jitter

`geom_jitter()` nudges each point by a small random amount so
overlapping points separate. It only changes how the plot looks - the
underlying data is unchanged.

```{r}
# Task: make the same avg_study_hours vs. gpa plot, but swap
#       geom_point() for geom_jitter()

```

## Adding a third variable: color

Mapping a variable to `color` inside `aes()` can reveal patterns a
single-color plot hides. We'll switch variables here: `avg_study_hours`
vs. `pea` (fitness score), colored by `is_ic` (IC athlete status) -
athlete status was built to affect *both* of these variables, so
coloring by it should visibly split the cloud into two groups.

```{r}
# Task: build a scatterplot of avg_study_hours (x) vs. pea (y), with
#       alpha = 0.2, colored by color = factor(is_ic) inside aes()

# Task: in a comment, compare this colored plot to what you'd see in
#       a single-color version - what shows up now that didn't before?

```

## Comparing groups: faceting

`facet_wrap(~ variable)` splits one plot into a grid of side-by-side
panels, one per group, instead of overlaying everything with color -
a second way to compare the same two groups.

```{r}
# Task: build the same avg_study_hours vs. pea scatterplot (no color
#       this time), then add facet_wrap(~ is_ic) instead

# Task: in a comment, which do you prefer for comparing IC athletes to
#       non-athletes here - the colored version or the faceted
#       version - and why?

```

## Putting it together

Combine what you've built: data, aesthetics (including a third
variable), geometry, a fix for overplotting, and labels, all in one
plot.

```{r}
# Task: build one scatterplot of avg_study_hours (x) vs. pea (y) that:
#       (1) uses alpha to handle overplotting, (2) colors points by
#       factor(is_ic), and (3) has a title and readable axis labels

```
